\Large\textbf{Advection versus diffusion in brain ventricular transport}

### 2025-04-08 paper/main.tex
update fig appendix, merge Miro edits

\newcommand{\miro}[1]{{\color{cyan}#1}}

@@

Assuming that viscous forces are dominating the ventricular CSF flow,
we model the flow by the Stokes equations.
The net forces of motile cilia acting on the CSF are modeled as a steady,
bidirectional tangential traction acting on parts of the ventricular wall
(\Crefpanels{fig:fig1}{d}), with \miro{no-penetration condition at the walls}.
Simulations show that this traction alone
leads to a partial compartmentalization of the ventricular system,
with large-scale vortex structures in the anterior, middle and
posterior ventricles (\Crefpanels{fig:fig1}{e}). The highest flow
speeds ($26.7$ \textmu m/s) occur in the vicinity of the dorsal cilia
in the middle ventricle. To model the cardiac pulsations, we impose a
pulsatile pressure difference between the anterior and posterior
boundaries (\Crefpanels{fig:fig1}{d, f}). The resulting flow
is pulsatile and directional (\Crefpanels{fig:fig1}{g}), and
in the absence of the cilia, the vortex structures vanish. The
velocity magnitude peaks at $4.8$ \textmu m/s. CSF initially flows
rostrocaudally before changing direction to caudorostral in the middle
of the cardiac cycle.

@@

Removing motile cilia delayed transport within
the middle ventricle and towards the anterior ventricle,
while the transport to the posterior ventricle was unaffected.
Photoconversion experiments with cilia mutant
zebrafish larva (\emph{smh}) aligned with these observations.
This further indicates that
transport from the middle to the posterior ventricle is
purely diffusive at this developmental stage,
while transport within the middle ventricle
depends on cilia motility.
Interestingly, Yoshida~\emph{et al.} observed a drastic reduction
in CSF mixing in specific regions of the brain
ventricles when simulating reduced cilia motility~\cite{Yoshida2022EffectVentricles}.
\fixme{Our experimental results provide not only confidence
in the predictive ability of our computational model,
but also quantify the impact of cilia motility
on brain ventricular transport.}

@@

The maximum stress of the cilia boundary condition was
$\vert\btau\vert_{\mathrm{max}} = 2.5 \ \mathrm{mPa} = \num{2.5e-3} \ \mathrm{N/m^2}$.
In a model of CSF flow in human brain ventricles,
Siyahhan~\emph{et al.} modeled ciliary motion with
a force increasing linearly from
the ventricular wall to a maximum force density of
$f_{\mathrm{max}} = 526 \ \mathrm{N/m^3}$ at
a distance 15 \textmu m (the cilium length)
away~\cite{Siyahhan2014FlowVentricles}.
In a cilia array of thickness $\delta = 15$ \textmu m,
the maximum stress in our model would correspond to a maximum force density of 
\begin{equation*}
    f_c = \frac{\vert\btau\vert_{\mathrm{max}}}{\delta}
        = \frac{\num{2.5e-3} \ \mathrm{N/m^2}}{\num{15e-6} \ \mathrm{m}}
        = 169 \ \mathrm{N/m^3}.
\end{equation*}
With $\delta=5$ \textmu m (the typical cilium length in zebrafish brain ventricles
\cite{Olstad2019CiliaryDevelopment,
Salman2022ComputationalEmbryo, Ringers2023NovelEpithelia}) 
our cilia stress corresponds to a maximum force density of
$f_c=507 \ \mathrm{N/m^3}$.
For further comparison, Thouvenin~\emph{et al.}
modeled two-dimensional cilia-mediated 
flow in the central canal of zebrafish
embryos using a force density of 4000 $\mathrm{N/m^2}$\cite{Thouvenin2020OriginCanal}.
Averaging this value over the reported central canal diameter 8.9 \textmu m
yields a maximum force of 449 $\mathrm{N/m^3}$.
In light of the previous finding that motile cilia
align by response to shear stress magnitudes
of flow,~\cite{Pellicciotta2020Cilia}
it is interesting to note the above similarity 
in cilia force estimates.

@@

Wholemount in vivo live imaging and photoconversion
experiments were performed with zebrafish larvae at the 2 dpf stage obtained from inbreeding of heterozygous
\emph{smh\textsuperscript{+/-};Tg(gfap:Gal4FF);Tg(5xUAS:Signal-Dendra2})
adult animals, as described by D'Gama~\emph{et al.}~\cite{DGama2024Cilia-mediatedBrain}. \miro{$<<$ undef ref here}
Controls consisted of either wild-type (\emph{smh}\textsuperscript{+/+})
or \emph{smh}\textsuperscript{+/-} heterozygous zebrafish from the same breeding.
The mutants were identified based on their curved body~\cite{Jau-NianChen1997Left-rightZebrafish}.

@@

\subsection{Numerical approximation of the Stokes equations}
The Stokes equations (\Cref{eq:stokes_eqs}) are discretized with a finite element method
in space, while in time, we employ a first-order implicit Euler discretization~\cite{butcher2016numerical}.
We set the timestep size $\Delta t = 0.023$ s such that 20 timesteps represent one cardiac cycle.
For the spatial discretization, we use piecewise linear Brezzi-Douglas-Marini (BDM)
elements for the velocity.~\cite{Brezzi1985TwoProblems, Nedelec1986AIR3}
The BDM elements degrees of freedom are associated with integral moments of the
velocity component normal to the facets of elements,
thus we enforce the impermeability condition $\uu\cdot\nn=0$ strongly.
We impose the boundary conditions for both the tangential traction $\bsigpar$ and
the normal pressure $\Tilde{p}(t)$ weakly.
For the pressure, we use piecewise constant (zeroth order discontinuous Galerkin (DG)) elements.
The resulting total number of degrees of freedom at each time step is 1 009 874.
The element combination $\mathrm{BDM}_1$--$\mathrm{DG}_{0}$ yields a non-conforming
discretization scheme for the Stokes equations~\cite{Stenberg1989SomeEquations}. \miro{In particular,
  with BDM only the normal component of the velocity is continuous.}
Continuity in tangential velocities on interior facets
is weakly enforced with a penalty parameter $\gamma=10$~\cite{Hong2016AEquations}.
Since the divergence of the velocity function space is the pressure function space for this scheme,
the divergence-free condition of the velocity is satisfied exactly, ensuring mass conservation.~\cite{Boffi2008FiniteProblem}

\subsection{Numerical approximation of the advection-diffusion equation}\label{subsec:num_approx_adv_diff}
\miro{The advection-diffusion equation (\Cref{eq:adv_diff_strong}) is discretized in
space with quadratic DG elements using symmetric interior penalty DG method~\cite{Arnold1982AnElements}.
The scheme was chosen due to its favorable mass 
conservation properties}. At each timestep the total number of degrees of freedom 
is 1 415 120. In time, the equation is discretized with the second-order backward
difference scheme BDF2.~\cite{Volker2016}
To accurately approximate the solute concentrations even
in the presence of strong advection (high Péclet numbers),
we use an upwind scheme for the velocity when calculating the
advective flux~\cite{Patankar2018NumericalFlow}.
As for the flow equations, we use a timestep size of
$\Delta t = 0.023$ s. We simulate transport for 1900 cardiac cycles,
with a final time of simulation $T = 856$ s.
A DG penalty parameter $\alpha = 50$ was chosen to ensure stability of the method.

@@

To derive weak formulations of the continuous problem defined by the Stokes
equations (\Cref{eq:appendix_stokes_eqs}), we consider the function space\footnote{
To simplify the exposition we focus only on spatial regularity. In particular,
$\vv\in\mathbf{H}^1(\Omega)$ is understood as $\vv(\cdot, t)\in \mathbf{H}^1(\Omega)$ for all
$t>0$. Similar convention is applied also when discussing advection-diffusion equation in \Cref{subsec:adv_diff_eq_weak_form}.
}
\begin{align*}
    \VV &= \bigg\{\vv \ : \ \vv\in\mathbf{H}^1(\Omega),
    \vv\cdot\nn=0 \ \mathrm{on} \ \Gamma\setminus\Gp\bigg\},
\end{align*}
for the velocity and $Q=L^2(\Omega)$ for the pressure.
Here $\Gp$ is the anterior/posterior boundary where we impose a normal pressure
boundary condition for the model versions that consider fluid motion induced by
cardiac pulsations. Bold-faced characters for a space denotes a $d$--dimensional vector space,
where $d$ is the spatial dimension of $\Omega$. For example,
$\mathbf{H}^1(\Omega) = [H^1(\Omega)]^3$ if $\Omega$ is three-dimensional.

@@

\miro{The elements of $\VV_h$ only ensure by construction the continuity of normal components of $\uu_h$.
Continuity of the tangential components of $\uu_h$ on interior facets of
the mesh is enforced weakly by additional stabilization to 
the discrete weak form (\Cref{eq:no_stabilization_abstract_discrete_weak_form_modelC}).
Specifically}, we apply the stabilization outlined by Hong~\emph{et al.}~\cite{Hong2016AEquations}.
First, we define the average and jump operators, respectively, as
\begin{equation}
    \avg{\uu} = \frac{\uu^+\cdot\nn^+ - \uu^-\cdot\nn^-}{2}
    \qquad\mathrm{and}\qquad \jump{\uu} = \uu^+ - \uu^-,
    \label{eq:operators_BDM}
\end{equation}
where $\uu$ is a first-order tensor and 
$\nn$ is the normal vector on a facet separating two cells,
and the plus and minus signs denote variables in each of the two cells.
The normal vector is chosen such that it points towards the exterior of
its corresponding cell. On the boundary, the operators are defined as
$\avg{\uu} = \uu\cdot\nn$ and $\jump{\uu} = \uu$,
with $\nn$ pointing to the exterior of the domain. Note that average operator
defined in~\Cref{eq:operators_BDM} reduces the rank of the tensor $\uu$ by one,
whereas the jump operator preserves rank. This extends to tensors of higher order.
We now define and add the bilinear form
\begin{align*}
    \mathcal{A}_{\mathrm{stab}}(\uu_h, \vv_h) &=
    \sum_{f\in \mathcal{F}}\int_f{2\mu\frac{\gamma}{\avg{h}}\jump{P_{\nn}(\uu_h)}
    \cdot\jump{P_{\nn}(\vv_h)}}\,\mathrm ds\\
    &-\sum_{f\in \mathcal{F}}\int_f{2\mu\avg{\beps(\uu_h)}\cdot\jump{\vv_h}}\,\mathrm ds
    -\sum_{f\in \mathcal{F}}\int_f{2\mu\avg{\beps(\vv_h)}\cdot\jump{\uu_h}}\,\mathrm ds
\end{align*}
to the discrete weak formulation (\Cref{eq:no_stabilization_abstract_discrete_weak_form_modelC}),
where $\mathcal{F}$ is the set of all interior facets of the computational mesh.
The first term of $\mathcal{A}_{\mathrm{stab}}$ penalizes jumps
in the tangential components of the velocity
across interior facets, where the penalty parameter $\gamma>0$ must be chosen large enough
to achieve stability. \miro{For appropriate convergence, the penalization term
includes a mesh-dependent scaling. In this work we choose it as the average of the cell diameter of neighboring mesh cells,
with $h$ defined as the maximal distance between two points in a cell~\cite{Nitsche1971UberSind}.}
The other two terms
of $\mathcal{A}_{\mathrm{stab}}$ are added to ensure consistency and symmetry of
the bilinear form, such that the problem is well-defined.

@@

Convergence rates reported in~\Cref{subsec:numerical_verification} were
approximated in the following way. Define the error $e_k$ for a given mesh
resolution $h_k$. Based on the assumption that the error scales as
$e_k \approx C{h_k}^\eta$, where $C$ is a constant independent of $h_k$,
the convergence order $\eta$ was approximated as
\begin{equation}
    \eta = \frac{\log{(e_k/e_{k+1}})}{\log{(h_k/h_{k+1}})}.
    \label{eq:convergence_order_estimate}
\end{equation}
\miro{
We define the approximation of the infinity norm of a finite element function
$\phi = \phi_i\psi_i(\xx)$
as 
\begin{equation*}
    \normlinf{\phi} = \max_i{\vert \phi_i\vert}.
\end{equation*}
Here $\psi_i$ are the basis functions and $\phi_i$ are the degrees of freedom defined as point evaluations}

@@

For all meshes, we report the maximum discrete velocity magnitude $\normlinf{u_h}$, \miro{$u_h=\sqrt{\uu_h\cdot\uu_h}$}.
We also report $\normlinf{u_h}$ scaled by the total force $F=\intGc{\sqrt{\btau\cdot \btau}}$
applied on the cilia boundary $\Gc$. We chose this measure using $F$ as a scale because the regions
of the mesh marked as cilia-populated change slightly with every mesh refinement,
owing to the unstructured organization of the mesh cells.
These changes in the surface area over which the tangential traction is applied
result in changes to $\normlinf{u_h}$,
since the maximum velocity magnitude is a result of the total amount of force
$F$ applied on the cilia boundary.

@@

\section{Quantification of velocities in particle tracking}\label{sec:appC_particle_transport}
Particle tracking data was analyzed~\cite{Olstad2019CiliaryDevelopment}.
For four zebrafish, five recordings with a few particles for each fish were carried out.
The particles tracked were in the duct between the rhombencephalic ventricle and the
diencephalic ventricle. Frequency of image acquisition was 20.22 Hz.
The collected data constituted $x$ and $y$ positions of the particles at every time
instant of image acquisition. The velocities $u_x$ and $u_y$ were determined by
calculating the change in positions $\Delta x$ and $\Delta y$ relative to the change
in time $\Delta t$ between two subsequent images, $\Delta t$ being equal to the
frequency of acquisition.
Mean velocities $\overline{u}_x$~(\Crefpanels{fig:particle_tracking_data}{a}) and
$\overline{u}_y$~(\Crefpanels{fig:particle_tracking_data}{b}),
which serve as a quantification of particle drift, were calculated
by taking the difference between the last and first positions of a particle.

Particle movement varied over the different measurements. The mean of $\overline{u}_x$ was
0.26 \textmu m/s and the mean of $\overline{u}_y$ was 0.08 \textmu m/s.
The maximum mean velocities
in absolute values were 10.2 \textmu m/s and 2.4 \textmu m/s for
$\overline{u}_x$ and $\overline{u}_y$,
respectively. We used the cardiac-induced/no-cilia flow model to tune the
pressure boundary condition
amplitude $A$. The choice of $A=1.5$ mPa was used since: (i) it yields a
maximum velocity magnitude
of 4.8 \textmu m/s, which is close to the maximum observed in the
particle tracking data when
rejecting the outlier at around 10 \textmu m/s; (ii) it yields a time-averaged mean velocity
of 0.89 \textmu m/s, which is close to the mean observed in the particle tracking data.
\begin{figure}[H]
    \begin{minipage}[t]{0.49\textwidth}
        \includegraphics[width=\textwidth]{graphics/mean_velocities_x.png}
    \end{minipage}
    \begin{minipage}[t]{0.49\textwidth}
        \includegraphics[width=\textwidth]{graphics/mean_velocities_y.png}
    \end{minipage} 
    \caption{Particle tracking data. Mean velocities in
    the $x$ (\textbf{a.}) and $y$ (\textbf{b.}) directions of the particles tracked.
    The dashed horizontal line designates a value of zero.}\label{fig:particle_tracking_data}
\end{figure}

>>>>

@@

Assuming that viscous forces are dominating the ventricular CSF flow,
we model the flow by the Stokes equations.
The net forces of motile cilia acting on the CSF are modeled as a steady,
bidirectional tangential traction acting on parts of the ventricular wall
(\Crefpanels{fig:fig1}{d}), with an impermeability condition at the walls.
Simulations show that this traction alone
leads to a partial compartmentalization of the ventricular system,
with large-scale vortex structures in the anterior, middle and
posterior ventricles (\Crefpanels{fig:fig1}{e}). The highest flow
speeds ($26.7$ \textmu m/s) occur in the vicinity of the dorsal cilia
in the middle ventricle. To model the cardiac pulsations, we impose a
pulsatile pressure difference between the anterior and posterior
boundaries (\Crefpanels{fig:fig1}{d, f}). The resulting flow
is pulsatile and directional (\Crefpanels{fig:fig1}{g}), and
in the absence of the cilia, the vortex structures vanish. The
velocity magnitude peaks at $4.8$ \textmu m/s. CSF initially flows
rostrocaudally before changing direction to caudorostral in the middle
of the cardiac cycle.

@@

Removing motile cilia delayed transport within
the middle ventricle and towards the anterior ventricle,
while the transport to the posterior ventricle was unaffected.
Photoconversion experiments with cilia mutant
zebrafish larva (\emph{smh}) aligned with these observations.
These simulations and experiments further indicate that
transport from the middle to the posterior ventricle is
purely diffusive at this developmental stage,
while transport within the middle ventricle
depends on cilia motility.
Furthermore, our experimental results provide quantification
of the impact of cilia motility
on brain ventricular transport.
Interestingly, Yoshida~\emph{et al.} observed a drastic reduction
in CSF mixing in specific regions of the human brain
ventricles when simulating reduced cilia motility~\cite{Yoshida2022EffectVentricles},
an observation that compares well with our findings.

@@

The maximum stress of the cilia boundary condition was
$\vert\btau\vert_{\mathrm{max}} = 2.5 \ \mathrm{mPa} = \num{2.5e-3} \ \mathrm{N/m^2}$.
In a model of CSF flow in human brain ventricles,
Siyahhan~\emph{et al.} modeled ciliary motion with
a force increasing linearly from
the ventricular wall to a maximum force density of
$f_{\mathrm{max}} = 526 \ \mathrm{N/m^3}$ at
a distance 15 \textmu m (the cilium length)
away~\cite{Siyahhan2014FlowVentricles}.
Their simulation results suggest that cilia only contribute to
near-wall flow dynamics, which contrasts our results in which
the cilia induce large-scale flow. This is a consequence of the 
scale of the cilia and the difference in ventricular size
between humans and zebrafish.
In a cilia array of thickness $\delta = 15$ \textmu m,
the maximum stress in our model would correspond to a maximum force density of 
\begin{equation*}
    f_c = \frac{\vert\btau\vert_{\mathrm{max}}}{\delta}
        = \frac{\num{2.5e-3} \ \mathrm{N/m^2}}{\num{15e-6} \ \mathrm{m}}
        = 169 \ \mathrm{N/m^3}.
\end{equation*}
With $\delta=5$ \textmu m (the typical cilium length in zebrafish brain ventricles
\cite{Olstad2019CiliaryDevelopment,
Salman2022ComputationalEmbryo, Ringers2023NovelEpithelia}) 
our cilia stress corresponds to a maximum force density of
$f_c=507 \ \mathrm{N/m^3}$, similar to the value
Siyahhan~\emph{et al.} used~\cite{Siyahhan2014FlowVentricles}.
For further comparison, Thouvenin~\emph{et al.}
modeled two-dimensional cilia-mediated 
flow in the central canal of zebrafish
embryos using a force density of 4000 $\mathrm{N/m^2}$\cite{Thouvenin2020OriginCanal}.
Averaging this value over the reported central canal diameter 8.9 \textmu m
yields a maximum force of 449 $\mathrm{N/m^3}$.
In light of the previous finding that motile cilia
align by response to shear stress magnitudes
of flow,~\cite{Pellicciotta2020Cilia}
it is interesting to note the above similarity 
in cilia force estimates, despite the variety in application.

@@

Wholemount in vivo live imaging and photoconversion
experiments were performed with zebrafish larvae at the 2 dpf stage obtained from inbreeding of heterozygous
\emph{smh\textsuperscript{+/-};Tg(gfap:Gal4FF);Tg(5xUAS:Signal-Dendra2})
adult animals, as described by D'Gama~\emph{et al.}~\cite{DGama2025MotileBrain}.
Controls consisted of either wild-type (\emph{smh}\textsuperscript{+/+})
or \emph{smh}\textsuperscript{+/-} heterozygous zebrafish from the same breeding.
The mutants were identified based on their curved body~\cite{Jau-NianChen1997Left-rightZebrafish}.

@@

\subsection{Numerical approximation of the Stokes equations}
The Stokes equations (\Cref{eq:stokes_eqs}) are discretized with a finite element method
in space, while in time, we employ a first-order implicit Euler discretization~\cite{butcher2016numerical}.
We set the timestep size $\Delta t = 0.023$ s such that 20 timesteps represent one cardiac cycle.
For the spatial discretization, we use piecewise linear Brezzi-Douglas-Marini (BDM)
elements for the velocity.~\cite{Brezzi1985TwoProblems, Nedelec1986AIR3}
The BDM elements degrees of freedom are associated with integral moments of the
velocity component normal to the facets of elements,
thus we enforce the impermeability condition $\uu\cdot\nn=0$ strongly.
We impose the boundary conditions for both the tangential traction $\bsigpar$ and
the normal pressure $\Tilde{p}(t)$ weakly.
For the pressure, we use piecewise constant (zeroth order discontinuous Galerkin (DG)) elements.
The resulting total number of degrees of freedom at each time step is 1 009 874.
The element combination $\mathrm{BDM}_1$--$\mathrm{DG}_{0}$ yields a non-conforming
discretization scheme for the Stokes equations~\cite{Stenberg1989SomeEquations}. 
In particular, with BDM only the normal component of the velocity is continuous.
Continuity in tangential velocities on interior facets
is weakly enforced with a penalty parameter $\gamma=10$~\cite{Hong2016AEquations}.
Since the divergence of the velocity function space is the pressure function space for this scheme,
the divergence-free condition of the velocity is satisfied exactly, ensuring mass conservation.~\cite{Boffi2008FiniteProblem}

\subsection{Numerical approximation of the advection-diffusion equation}\label{subsec:num_approx_adv_diff}
The advection-diffusion equation (\Cref{eq:adv_diff_strong}) is discretized in
space with quadratic DG elements using a symmetric interior penalty DG method~\cite{Arnold1982AnElements}.
The scheme was chosen due to its favorable mass 
conservation properties. At each timestep the total number of degrees of freedom 
is 1 415 120. In time, the equation is discretized with the second-order backward
difference scheme BDF2.~\cite{Volker2016}
To accurately approximate the solute concentrations even
in the presence of strong advection (high Péclet numbers),
we use an upwind scheme for the velocity when calculating the
advective flux~\cite{Patankar2018NumericalFlow}.
As for the flow equations, we use a timestep size of
$\Delta t = 0.023$ s. We simulate transport for 1900 cardiac cycles,
with a final time of simulation $T = 856$ s.
A DG penalty parameter $\alpha = 50$ was chosen to ensure stability of the method.

@@

To derive weak formulations of the continuous problem defined by the Stokes
equations (\Cref{eq:appendix_stokes_eqs}), we consider the function space\footnote{
To simplify the exposition we focus only on spatial regularity. In particular,
$\vv\in\mathbf{H}^1(\Omega)$ is understood as $\vv(\cdot, t)\in \mathbf{H}^1(\Omega)$ for all
$t>0$. Similar convention is applied also when discussing the advection-diffusion equation in \Cref{subsec:adv_diff_eq_weak_form}.
}
\begin{align*}
    \VV &= \bigg\{\vv \ : \ \vv\in\mathbf{H}^1(\Omega),
    \vv\cdot\nn=0 \ \mathrm{on} \ \Gamma\setminus\Gp\bigg\},
\end{align*}
for the velocity and $Q=L^2(\Omega)$ for the pressure.
Here $\Gp$ is the anterior/posterior boundary where we impose a normal pressure
boundary condition for the model versions that consider fluid motion induced by
cardiac pulsations. Bold-faced characters for a space denotes a $d$--dimensional vector space,
where $d$ is the spatial dimension of $\Omega$. For example,
$\mathbf{H}^1(\Omega) = [H^1(\Omega)]^3$ if $\Omega$ is three-dimensional.

@@

The elements of $\VV_h$ only ensure by construction the continuity of normal components of $\uu_h$.
Continuity of the tangential components of $\uu_h$ on interior facets of
the mesh is enforced weakly by additional stabilization to 
the discrete weak form (\Cref{eq:no_stabilization_abstract_discrete_weak_form_modelC}).
Specifically, we apply the stabilization outlined by Hong~\emph{et al.}~\cite{Hong2016AEquations}.
First, we define the average and jump operators, respectively, as
\begin{equation}
    \avg{\uu} = \frac{\uu^+\cdot\nn^+ - \uu^-\cdot\nn^-}{2}
    \qquad\mathrm{and}\qquad \jump{\uu} = \uu^+ - \uu^-,
    \label{eq:operators_BDM}
\end{equation}
where $\uu$ is a first-order tensor and 
$\nn$ is the normal vector on a facet separating two cells,
and the plus and minus signs denote variables in each of the two cells.
The normal vector is chosen such that it points towards the exterior of
its corresponding cell. On the boundary, the operators are defined as
$\avg{\uu} = \uu\cdot\nn$ and $\jump{\uu} = \uu$,
with $\nn$ pointing to the exterior of the domain. Note that average operator
defined in~\Cref{eq:operators_BDM} reduces the rank of the tensor $\uu$ by one,
whereas the jump operator preserves rank. This extends to tensors of higher order.
We now define and add the bilinear form
\begin{align*}
    \mathcal{A}_{\mathrm{stab}}(\uu_h, \vv_h) &=
    \sum_{f\in \mathcal{F}}\int_f{2\mu\frac{\gamma}{\avg{h}}\jump{P_{\nn}(\uu_h)}
    \cdot\jump{P_{\nn}(\vv_h)}}\,\mathrm ds\\
    &-\sum_{f\in \mathcal{F}}\int_f{2\mu\avg{\beps(\uu_h)}\cdot\jump{\vv_h}}\,\mathrm ds
    -\sum_{f\in \mathcal{F}}\int_f{2\mu\avg{\beps(\vv_h)}\cdot\jump{\uu_h}}\,\mathrm ds
\end{align*}
to the discrete weak formulation (\Cref{eq:no_stabilization_abstract_discrete_weak_form_modelC}),
where $\mathcal{F}$ is the set of all interior facets of the computational mesh.
The first term of $\mathcal{A}_{\mathrm{stab}}$ penalizes jumps
in the tangential components of the velocity
across interior facets, where the penalty parameter $\gamma>0$ must be chosen large enough
to achieve stability. For appropriate convergence, the penalization term
includes a mesh-dependent scaling. In this work,
we choose it as the average of the cell diameter of neighboring mesh cells,
with $h$ defined as the maximal distance between two points in a cell~\cite{Nitsche1971UberSind}.
The other two terms
of $\mathcal{A}_{\mathrm{stab}}$ are added to ensure consistency and symmetry of
the bilinear form, such that the problem is well-defined.

@@

Convergence rates reported in~\Cref{subsec:numerical_verification} were
approximated in the following way. Define the error $e_k$ for a given mesh
resolution $h_k$. Based on the assumption that the error scales as
$e_k \approx C{h_k}^\eta$, where $C$ is a constant independent of $h_k$,
the convergence order $\eta$ was approximated as
\begin{equation}
    \eta = \frac{\log{(e_k/e_{k+1}})}{\log{(h_k/h_{k+1}})}.
    \label{eq:convergence_order_estimate}
\end{equation}

We define the approximation of the infinity norm of a finite element function
$\phi = \phi_i\psi_i(\xx)$
as 
\begin{equation*}
    \normlinf{\phi} = \max_i{\vert \phi_i\vert}.
\end{equation*}
Here $\psi_i$ are the basis functions and $\phi_i$ are the degrees of freedom defined as point evaluations.

@@

For all meshes, we report the maximum discrete velocity magnitude $\normlinf{u_h}$, with $u_h=\sqrt{\uu_h\cdot\uu_h}$.
We also report $\normlinf{u_h}$ scaled by the total force $F=\intGc{\sqrt{\btau\cdot \btau}}$
applied on the cilia boundary $\Gc$. We chose this measure using $F$ as a scale because the regions
of the mesh marked as cilia-populated change slightly with every mesh refinement,
owing to the unstructured organization of the mesh cells.
These changes in the surface area over which the tangential traction is applied
result in changes to $\normlinf{u_h}$,
since the maximum velocity magnitude is a result of the total amount of force
$F$ applied on the cilia boundary.

@@

\section{Quantification of velocities in particle tracking}\label{sec:appC_particle_transport}
Particle tracking data was analyzed~\cite{Olstad2019CiliaryDevelopment}.
For four zebrafish, five recordings with a few particles for each fish were carried out.
The particles tracked were in the duct between the rhombencephalic ventricle and the
diencephalic ventricle. Frequency of image acquisition was 20.22 Hz.
The collected data constituted $x$ (rostrocaudal axis) and $z$ (dorsoventral axis) positions of the particles at every time
instant of image acquisition. The velocities $u_x$ and $u_z$ were determined by
calculating the change in positions $\Delta x$ and $\Delta z$ relative to the change
in time $\Delta t$ between two subsequent images, $\Delta t$ being equal to the
frequency of acquisition.
Mean velocities $\overline{u}_x$~(\Crefpanels{fig:particle_tracking_data}{a}) and
$\overline{u}_z$~(\Crefpanels{fig:particle_tracking_data}{b}),
which serve as a quantification of particle drift, were calculated
by taking the difference between the last and first positions of a particle.

Particle movement varied over the different measurements. The mean of $\overline{u}_x$ was
0.26 \textmu m/s and the mean of $\overline{u}_z$ was 0.08 \textmu m/s.
The maximum mean velocities
in absolute values were 10.2 \textmu m/s and 2.4 \textmu m/s for
$\overline{u}_x$ and $\overline{u}_z$,
respectively. We used the cardiac-induced/no-cilia flow model to tune the
pressure boundary condition
amplitude $A$. The choice of $A=1.5$ mPa was used since: (i) it yields a
maximum velocity magnitude
of 4.8 \textmu m/s, which is close to the maximum observed in the
particle tracking data when
rejecting the outlier at around 10 \textmu m/s; (ii) it yields a time-averaged mean velocity
of 0.89 \textmu m/s, which is close to the mean observed in the particle tracking data.
\begin{figure}[H]
    \centering
    \includegraphics[width=\textwidth]{graphics/figure7.png}
    \caption{Particle tracking data. Mean velocities in
    the $x$ (\textbf{a}) and $z$ (\textbf{b})
    directions of the particles tracked.
    The dashed horizontal lines designate zero.}\label{fig:particle_tracking_data}
\end{figure}
